RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-05-01 Response to reviewers.txt

######## EDITOR ########

===== COMMENT #1
Please ensure also that your manuscript shows clearly how your work relates to relevant works as they may have been published in the Requirements Engineering or presented at the Requirements Engineering series of conferences.

===== REPLY #1
Currently 6 of the 44 references in the article come from either REJ or the RE conference:
- Y. Elrakaiby, A. Ferrari, P. Spoletini, S. Gnesi, and B. Nuseibeh. Using argumentation to explain ambiguity in requirements elici- tation interviews. In Proceedings of the 25th IEEE International Requirements Engineering Conference (RE 2017), pages 51–60. IEEE Press, 2017.
- I. Jureta, S. Faulkner, and P. Schobbens. Clear justification of mod- eling decisions for goal-oriented requirements engineering. Re- quirements Engineering, 13(2):87–115, May 2008.
- Y. Liu, Y. Su, X. Yin, and G. Mussbacher. Combined goal and feature model reasoning with the user requirements notation and jucmnav. In Requirements Engineering Conference (RE), 2014 IEEE 22nd International, pages 321–322. IEEE, 2014.
- P. K. Murukannaiah, A. K. Kalia, P. R. Telangy, and M. P. Singh. Resolving goal conflicts via argumentation-based analysis of com- peting hypotheses. In Proceedings of the 23rd International Re- quirements Engineering Conference (RE 2015), pages 156–165. IEEE, 2015.
- M. van Zee, F. Bex, and S. Ghanavati. Rationalization of Goal Models in GRL using Formal Argumentation. In Proceedings of the 23rd International Requirements Engineering Conference (RE’15), pages 220–225. IEEE Press, August 2015
- E. S. K. Yu. Towards Modeling and Reasoning Support for Early-Phase Requirements Engineering. In Proceedings of the 3rd IEEE International Symposium on Requirements Engineering, pages 226–235, 1997.

We have searched through the journal and conference database for more references, and we found 4 more references that are closely related to our work:
- S. R. Smith. Constructing and revising formal arguments in re- quirements engineering. Requir. Eng., 2(4):199–216, Dec. 1997.
- J.Lockerbie,N.A.M.Maiden,J.Engmann,D.Randall,S.Jones, and D. Bush. Exploring the impact of software requirements on system-wide goals: a method using satisfaction arguments and i* goal modelling. Requirements Engineering, 17(3):227–254, 2012.
- J. Horkoff and E. Yu. Interactive goal model analysis for early requirements engineering. Requirements Engineering, 21(1):29– 61, 2016.
- J. Hassine and D. Amyot. A questionnaire-based survey method- ology for systematically validating goal-oriented models. Require- ments Engineering, 21(2):285–308, 2016.

This means we currently have 10 references from the Requirements Engineering Journal or presented at the Requirements Engineering series of conferences.

######## REVIEWER #1 ########

We thank the reviewer for the insightful comments. Especially the comment about the missing validation was useful since it motivated us to perform an small-scale in-house user study, which allowed us to strengthen our claims for the usefulness and usability of our framework. We’ve replied to each separate comment below.

===== COMMENT #1
The integrated meta model looks reasonable but I have some doubts about the inclusion of "critical question" in the schema. Critical questions are something that are used to identify arguments and attack links, but  I don't see them as a concept to be included in the meta model. They are part of the methodology used to build the goal model rather concepts of the language.

===== REPLY #1
We thank the reviewer for pointing this out. We indeed believe critical questions should not be part of the metamodel, and we have removed them accordingly. We initially had put them into the metamodel because they play an important part in the implementation, but they are by themselves not part of the RationalGRL language. We also have updated the text accordingly.

===== COMMENT #2
The proposed algorithms are very simple and they do not add much to the overall presentation, I think they can be summarized in the section describing the tool.

===== REPLY #2
We agree that many algorithms are very much the same, and we have therefore shortened the sections containing these algorithms: sections 5.4.1 to 5.4.3 now all contain one example of the specific type of algorithm instead of two or more.

===== COMMENT #3
The methodology and the tools looks quite preliminary and without a real validation of the approach. I think this is the weakest part of the paper. Authors should demonstrate the applicability, scalability and usefulness of their approach in the real word. The tool and the modeling language should be used and evaluated by practitioners while they cope with real requirements engineering problems and complexity issues.

===== REPLY #3
TODO: WRITE WHEN ANALYSIS IS DONE

######## REVIEWER #2 ########

We thank the reviewer for reading our article in detail, and providing many insightful and useful comments. We’ve replied to each separate comment below.

===== COMMENT #1
I think that the relatively simple argument structures are appealing in the sense that the overhead for practical application may be tolerable. However, this remains to be shown yet in an industrial environment.

===== REPLY #1
TODO: WRITE WHEN ANALYSIS IS DONE

===== COMMENT #2
The statement on actually using a reduced GRL should be made in the Abstract and Introduction already, not just in a footnote later.

===== REPLY #2
We have included in the abstract and introduction that we use a simplified version. Note that all the core elements of GRL are included but, for example, we have included only one level of contribution strength, where GRL contains different levels of positive and negative contribution. This simplification does not in any way influence the case study or any of the other results. Furthermore, extending the GRL part of RationalGRL to include the full set of GRL elements would not change the essence of our formalism.

===== COMMENT #3
I think that the paper should compare the proposed argumentation structure at least with the beliefs in GRL. In general, how does this argumentation approach really differ from one based on beliefs and, in particular, belief revision? Assume that an argument is a belief of some stakeholder revised by another argument, which is a belief of some other stakeholder?

===== REPLY #3
%FB: **Floris to write on GRL beliefs in 7.1**

===== COMMENT #4
How would you compare this specific approach with the old gIBIS?

===== REPLY #4
%FB: **Floris to write on gIBIS in 7.1**

===== COMMENT #5
Please discuss why you are trying to have detailed, while simple, rationale, although the recent trend in DR is otherwise?

===== REPLY #5
%FB: **Floris to write on this in 7.1**

===== COMMENT #6
I wonder about the formalization in propositional logic. Why is it not used for automated reasoning (at least partly instead of the specific algorithms)?

===== REPLY #6
Propositional logic in itself does not contain specific algorithms for, for example, INTRO, REPLACE and DISABLE. Furthermore, the algorithms also introduce arguments and attacks, formal notions of which are not in basic propositional logic. The idea is that the formalization is needed for the algorithms, but that just providing a formal semantics of GRL does not directly tell you how to compute/reason with GRL models.

===== COMMENT #7
I am also missing a discussion about the consequences of this binary "nature" of argumentation with respect to argumentation in reality.

===== REPLY #7
%FB: **Floris to write on this in 5.2**

===== COMMENT #8
Regarding terminology, I wonder about "valid" GRL models here, which seem to mean that certain criteria are fulfilled. Validation would mean that a goal model makes sense in the real world as having the "right" goals and relationships between them.

===== REPLY #8
We agree that the term “valid” is confusing in this respect, since we do not want to express that the goal model makes sense in the real world. Instead, we’d like to express that certain criteria are fulfilled. Therefore, we’ve renamed “valid” to “consistent”. 
TODO(marcvanzee): Implement this renaming in the article.

===== COMMENT #9
"Verification" in 6.2 appears to stand for evaluation. This cannot be "verified".

===== REPLY #9
We’ve renamed “verification” to “demonstration” and slightly reworded the sentence.

===== COMMENT #10
I also do not think that Figure 16 shows a "methodology". It does not even look like a "method".

===== REPLY #10
**Sepideh**

===== COMMENT #11
please reconsider the very first sentences in the Abstract and Introduction each, especially the latter statement on RE

===== REPLY #11
**Sepideh**

===== COMMENT #12
Finally, you should run a spell checker, have a careful proof reading of grammar, and make sure that there is no unresolved reference to a figure in a submitted manuscript.

===== REPLY #11
We’ve ran a spell-checker and proof-read the paper carefully. There was one unresolved reference to a figure, which a corrected. Thank you for pointing this out.

### 2018-05-01 discussion.tex
Revision for final version

One other example of argumentation in RE concerns the use of argumentation in requirements elicitation. Ionita \emph{et al.}~\cite{ionita2014argumentation} propose a simple argumentation dialogue game in which risk assessors try to attack each other's arguments for certain risks attached to a system design. Dung's semantics \cite{Dung1995} are then used to determine the risks that are still valid, and those that have been successfully rejected. Elrakaiby \emph{et al.}~\cite{ElrakaibyFSGN17} use argumentation to explain ambiguity. They define when a statement by a client who is being interviewed about the requirements of a system presents an inconsistency or an insufficiency. An inconsistency can be either with the client's previous statements or with the requirement engineer's beliefs. An insufficiency occurs when an analyst needs more information from the client to accept a client's statement. The inconsistencies are then captured as mutually attacking arguments, and the insufficiencies as arguments against the original statements. These counterarguments say that, for example, the functionality expressed in the statement cannot be realized or is irrelevant. Elrakaiby \emph{et al.} coded the data from 34 requirement elicitation interviews. They identified 39 inconsistencies (i.e. at least two arguments that mutually attack) and 29 insufficiencies (i.e., at least one argument attacking another).

>>>>

One other example of argumentation in RE concerns the use of argumentation in requirements elicitation. Ionita \emph{et al.}~\cite{ionita2014argumentation} propose a simple argumentation dialogue game in which risk assessors try to attack each other's arguments for certain risks attached to a system design. Dung's semantics \cite{Dung1995} are then used to determine the risks that are still valid, and those that have been successfully rejected. Elrakaiby \emph{et al.}~\cite{ElrakaibyFSGN17} use argumentation to explain ambiguity. They define when a statement by a client who is being interviewed about the requirements of a system presents an inconsistency or an insufficiency. An inconsistency can be either with the client's previous statements or with the requirement engineer's beliefs. An insufficiency occurs when an analyst needs more information from the client to accept a client's statement.

Hassin and Amyot~\cite{hassine2016questionnaire} analyze the correctness of GRL models by performing statistical analysis on generated questionnaires. This allows early detection of potential conflicts between stakeholders of the system. Instead of questionnaires, we use arguments and make them explicitly traceable to elements of the goal model. Horkoff and Yu~\cite{horkoff2016interactive} introduce analysis procedures for the $i*$ goal-oriented framework, allowing users to ask questions such as ``what if?'', or ``are certain goals achievable? how? or why not?''. They formally defined their procedures and describe a framework implementation. Also in $i*$, Lockerbie \emph{et al.}~\cite{lockerbie2012exploring} describe procedures that analysts follow to use satisfaction arguments. They demonstrate their results in two case studies, providing evidence towards the claim that the integration of $i*$ and satisfaction arguments is useful and usable. About 30 years ago, Smith~\cite{Smith:1997:CRF:2737426.2737493} already proposed a framework to develop arguments for the confidences that a software-based system will be successful in a given environment. He used a simple temporal logic with compositional reasoning, allowing arguments to be constructed, and revised.

### 2018-05-01 methodology.tex
Revision for final version

In previous sections, we have shown how the RationalGRL framework can capture stakeholder discussions, and how interactions between two types of reasoning, practical reasoning and goal modeling, leads to two interlinked models, RationalGRL and GRL models. The previous section contained a formalization of RationalGRL, which forms the starting point of this section. In this section we clarify how practitioners can actually use the RationalGRL framework by proposing a methodology (\textbf{requirement 4}) and discussing a prototype RationalGRL tool (\textbf{requirement 5}).

@@

We propose the methodology shown in Figure~\ref{fig:rationalgrl-methodology} to develop a RationalGRL model, an earlier version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. %Here we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

\textbf{(1) Instantiate Argument Schemes (AS)} -- We start with the list of argument schemes (Table~\ref{table:argument-schemes}). Whilst discussing the requirements, we select schemes from the list and instantiate them to form arguments for GRL model elements. In this way we build or modify the GRL model by introducing new GRL elements (\textsf{INTRO}). Note that it is also possible to start modifying an existing GRL model which was not built using the RationalGRL methodology: each GRL element corresponds to an argument (i.e. an instantiated scheme), so it is possible to instantiate argument schemes based on an existing GRL model.

\textbf{(2) Answer Critical Questions (CQs)} -- After building or modifying the initial GRL model, we ask the relevant critical questions. Because each element in the GRL model corresponds to an instantiated scheme, we can look at Table~\ref{table:argument-schemes}) to see which questions are relevant given our GRL model.

\textbf{(3) Decide on Intentional Elements and their Relationships} -- By answering a critical question, one of the \textsf{INTRO}, \textsf{DISABLE} or \textsf{REPLACE} are performed on the GRL model. Any of these operations impact the arguments and corresponding GRL intentional elements, modifying the initial GRL model into a RationalGRL model. After these modifications, we can keep on asking critical questions (e.g. about elements that were introduced by previously answering a critical question) until we are satisfied with our model.

@@

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of integrated tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent verification to ensure our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations. The tool can be accessed from:

@@

GRL is implemented in the open-source tool jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2.000 commits made by over 40 contributors, representing over 200.000 lines of code. jUCMNav is used actively in research, and the tool has been extended in many directions over the past few years, such as for modeling and analyzing security misuse cases \cite{}, supporting activity theory \cite{}, Web-based goal modeling \cite{}, and enterprise architecture modeling \cite{}. \todo{Floris}{Marc}{references need to be added here!} Furthermore, jUCMNav includes Use Case modelling and contains various GRL evaluation algorithms with which the satisfiability of goals can be determined.

The jUCMNav tool is implemented as an Eclipse plugin. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is more suited for current purposes. For example, it is our experience is that it can take significant time to set up the Eclipse environment and install the jUCMNav plugin, which hampers quick uptake by practitioners and also makes setting up empirical evaluations more difficult. Since RationalGRL does not make use of the many features of jUCMNav, we believe a light-weight version is sufficient for our current aims, and may benefit the community in general.

@@

\paragraph{IE and links details pane} Figure~\ref{fig:tool:overview} shows the details pane for the goal `Generate cars'. The user can change the name of the element by changing the name and clicking `Rename', which will update the naming history on the bottom of the pane. The naming history is (simplified) implemenetation of the \textsf{REPLACE} operation on IEs. If the IE has decomposition links, the user can change the decomposition type as well. Each IE and link has a set of associated critical questions, which can be answered in the details pane by click on the button next to the question.

\paragraph{Critical question details pane} The details panel can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal element ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

@@

\paragraph{Argumentation semantics} The RationalGRL tool computes the acceptability of arguments on the fly. In the example of Figure~\ref{fig:tool:cqeffect}, the original (argument for) softgoal ``Easy to use'' is rejected because it's only attacker ``CQ11 - Redundant'' is accepted. However, if we then attack ``CQ11 - Redundant'' with a new argument ``Necessary'', the original argument for ``Easy to use'' is again accepted because its only attacker is rejected  (Figure~\ref{fig:tool:argument}). Note that when computing the acceptability of arguments, the RationalGRL tool makes the assumption that there are no attack cycles in the model and that hence there is one unique preferred extension (cf. Section~\ref{sect:formalframework:rationalgrl}) -- if the user creates an attack cycle, an error message is shown.

@@

\paragraph{Limitations}
For usability purposes, the \textsf{REPLACE} operation has been implemented differently than in our formal framework (Section~\ref{sect:formalframework:disable}). In the formal framework, a \textsf{REPLACE} introduces a new argument that replaces (and therefore attacks) all previous arguments with the same identifier. Including \textsf{REPLACE} like this in the tool would mean that, when the name of an element is changed, we would need to render all the previous versions of the element on the canvas of the tool. Because this would quickly become very cumbersome, we decided to implement a naming history for each element. For example, in Figure~\ref{fig:tool:overview}, the goal ``Generate cars'' was previously named ``Add cars''. In this way, a history is kept of how an element name was clarified without cluttering the model canvas. Similarly, the decomposition type can be changed in the details pane (Figure~\ref{fig:tool:overview}) without introducing new elements that attack the original ones (cf. Figure~\ref{fig:example-small3}, where the old decomposition type $AND$ of ``Generate traffic'' is shown as an explicit argument).

>>>>

In previous sections, we have shown how the RationalGRL framework can capture stakeholders discussions, and how interactions between two types of reasoning, practical reasoning and goal modeling, leads to two interlinked models, RationalGRL and GRL models. The previous section contained a formalization of RationalGRL, which forms the starting point of this section. In this section we clarify how practitioners can actually use the RationalGRL framework by proposing a methodology (\textbf{requirement 4}) and discussing a prototype RationalGRL tool (\textbf{requirement 5}).

@@

We propose the methodology shown in Figure~\ref{fig:rationalgrl-methodology} to develop a RationalGRL model, a version of which was presented at the 2017 iStar workshop \cite{ghanavatiMethodology}. Here we assume that the initial GRL models have been created based on the requirements specification documents and the discussions of the stakeholders. The rest of the steps are as follows:

\textbf{(1) Instantiate Argument Schemes (AS)} -- In this step, we start from the list of arguments schemes (Table~\ref{table:argument-schemes}). Whilst discussing the requirements, we select schemes from the list and instantiate them to form arguments for GRL IEs or links. In this way we build the GRL model by introducing new GRL elements (\textsf{INTRO}). For example, an argument scheme can be "Goal \emph{G} contributes to softgoal \emph{S}". When an argument scheme is instantiated, it corresponds to an argument for or against part of a goal model.

\textbf{(2) Answer Critical Questions (CQs)} -- After building or modifying the initial GRL model, we ask the relevant critical questions. Since each element in the GRL model corresponds to an instantiated scheme, we can look at Table~\ref{table:argument-schemes}) to see which questions are relevant given our GRL model. For example, for the argument scheme, "Goal \emph{G} contributes to softgoal \emph{S}", there are two critical questions as follows:  \emph{Does the goal contributes to the softgoal?} and \emph{Does the goal contributes to some other softgoals?}. When the analyst answers  a critical question, a new argument scheme may be instantiated. This is done in step (3) when the operation INTRO is executed.

\textbf{(3) Decide on Intentional Elements and their Relationships} -- By answering a critical question, one of the three operations (\textsf{INTRO}, \textsf{DISABLE} or \textsf{REPLACE}) is applied to the GRL model. Any of these operations impact the arguments and corresponding GRL intentional elements, modifying the initial GRL model into a RationalGRL model. Recall that \textsf{INTRO} means that 
a new argument scheme is created. That means, the current argument related to the critical question does not get attacked.  In the case of \textsf{DISABLE}, the intentional element or its related links are disabled in the model. \textsf{REPLACE} introduces a new argument and attacks the original argument at the same time. This means that the original element of the argument scheme is replaced with a new one.

@@

The final requirement of our framework is that it has tool support (\textbf{requirement 5}). This is important for various reasons: i) although there are various approaches attempting to combine goal modeling with argumentation, we are not aware of existing tool support (see Section~\ref{sect:discussion}), ii) it allows us to do user tests in future research, exploring the difference between GRL and RationalGRL, iii) tool support is an excellent demonstration to show that our formal framework and algorithms actually work. In this section we briefly highlight some features of the tool, and we explain some of its current limitations. The tool can be accessed from:

@@

GRL is implemented in the open-source tool jUCMNav~\cite{jUCMNav}, which is actively maintained\footnote{\url{http://jucmnav.softwareengineering.ca/foswiki/ProjetSEG}}. jUCMNav currently has over 2,000 commits made by over 40 contributors, representing over 200,000 lines of code. jUCMNav is used actively in research, and the tool has been extended in many directions over the past few years, such as for modeling and analyzing security misuse cases~\cite{daramola2012ontology}, supporting activity theory \cite{georg2015synergy}, combined goal and feature model reasoning~\cite{liu2014combined}, and enterprise architecture modeling \cite{marosin-etal:caise2016}. Furthermore, jUCMNav includes Use Case modeling and contains various GRL evaluation algorithms with which the satisfiability of goals can be determined.

The jUCMNav tool is implemented as an Eclipse plugin. While we aim to integrate our framework into jUCMNav in the future, a web-based version of RationalGRL (and GRL) is more suited for current purposes. Since RationalGRL does not make use of the many features of jUCMNav, we believe a light-weight version is sufficient for our current aims, and may benefit the community in general. Our tool is thus not meant to replace jUCMNav. For simplicity we have a web-based version, and it has the functionality to export to jUCMNav.

@@

\paragraph{IE and links details pane} Figure~\ref{fig:tool:overview} shows the details pane for the goal `Generate cars'. The user can change the name of the IE by changing the name and clicking `Rename', which will update the naming history on the bottom of the pane. The naming history is (simplified) implementation of the \textsf{REPLACE} operation on IEs, since it is a very simple form of \emph{clarification}. The main difference with jUCMNav is that we have a renaming history, so the user can see which names the IE had prior to the current one. If the IE has decomposition links, the user can change the decomposition type as well. Each IE and link has a set of associated critical questions, which can be answered in the details pane by click on the button next to the question.

\paragraph{Critical question details pane} The details panel can be used to answer the critical questions of the RationalGRL framework (Table~\ref{table:argument-schemes}): Figure~\ref{fig:tool:overview}, for example, shows two critical questions for the goal IE ``Generate cars'', namely ``Can the goal be realized?'' (CQ3) and ``Is the goal relevant/useful?'' (CQ11). As an example, consider Figure~\ref{fig:tool:cqdetails}, where the softgoal ``Easy to use'' is questioned with the relevancy question (CQ11). It is possible to select the answer and provide an explanation for the answer.

@@

\paragraph{Argumentation semantics} The RationalGRL tool computes the acceptability of arguments on the fly. In the example of Figure~\ref{fig:tool:cqeffect}, the original (argument for) softgoal ``Easy to use'' is rejected because its only attacker ``CQ11 - Redundant'' is accepted. However, if we then attack ``CQ11 - Redundant'' with a new argument ``Necessary'', the original argument for ``Easy to use'' is again accepted because its only attacker is rejected  (Figure~\ref{fig:tool:argument}). Note that when computing the acceptability of arguments, the RationalGRL tool makes the assumption that there are no attack cycles in the model and that hence there is one unique preferred extension (cf. Section~\ref{sect:formalframework:rationalgrl}) -- if the user creates an attack cycle, an error message is shown.

@@

\paragraph{Limitations}
For usability purposes, the \textsf{REPLACE} operation has been implemented differently than in our formal framework (Section~\ref{sect:formalframework:disable}). In the formal framework, a \textsf{REPLACE} introduces a new argument that replaces (and therefore attacks) all previous arguments with the same identifier. Including \textsf{REPLACE} like this in the tool would mean that, when the name of an element is changed, we would need to render all the previous versions of the element on the canvas of the tool. Since this would quickly become very cumbersome, we decided to implement a naming history for each element. For example, in Figure~\ref{fig:tool:overview}, the goal ``Generate cars'' was previously named ``Add cars''. In this way, a history is kept of how an element name was clarified without cluttering the model canvas. This feature is currently very limited and can be extended in various was, for instance by changing the IE type. Similarly, the decomposition type can be changed in the details pane (Figure~\ref{fig:tool:overview}) without introducing new elements that attack the original ones (cf. Figure~\ref{fig:example-small3}, where the old decomposition type $AND$ of ``Generate traffic'' is shown as an explicit argument).
